---
title: "Stats207 Final Project Code"
output: html_document
date: "2024-06-02"
---

```{r}
data = read.csv("data.csv")
str(data)
# interested in daily changes: extracted hourly values at hour = 12
data = data[data$hour == 12, ]
```

```{r}
# hourly pm2.5
y = ts(data$pm2.5) 

# filling NAs 
# method1: forward then backward filling
library(zoo)
y_forward_filled = na.locf(y, na.rm = FALSE)
y_backward_filled = na.locf(y_forward_filled, fromLast = TRUE)
y = y_backward_filled

# method2: linear interpolation
library(forecast)
y = na.interp(y)

# both methods give nearly the same results
```

original data
```{r}
plot(y) # does not look stationary
acf(y, na.action = na.pass, lag.max = 100)
pacf(y, na.action = na.pass,lag.max = 100)
```

## Linear models

### polynomial + seasonal components

```{r}
library(lmtest) 
library(car)
library(stats)  
```

Include a time index for polynomial terms and generate seasonal dummy variables:

```{r}
data$date = as.Date(paste(data$year, data$month, data$day, sep="-"))
data$index = as.numeric(data$date - min(data$date))
data$index2 = data$index^2
data$index3 = data$index^3
data$month = as.factor(data$month)
data$month2 = data$month^2

train = data[1:1460, ] #80%
test = data[1461:nrow(data),] #20%
y_train = ts(train$pm2.5) 
y_test = ts(test$pm2.5) 
y_train = na.interp(y_train)
y_test = na.interp(y_test)
```

Fit linear regression model:

```{r}
model = lm(y_train ~ month - 1, data=train)
summary(model)
```

Diagnostics of first linear model: check for multicollinearity, heteroscedasticity, and model fit:

```{r}
# vif(model)

par(mfrow=c(2,2))
plot(model)

bptest(model)  # Breusch-Pagan test for heteroscedasticity
dwtest(model)  # Durbin-Watson test for autocorrelation

```


### polynomial + seasonal components + other features

Convert wind direction into categorical variable with one-hot enccoding:
```{r}
train = cbind(train, model.matrix(~ cbwd - 1, data=train))
test = cbind(test, model.matrix(~ cbwd - 1, data=test))
```

Fit model:
```{r}
full_model = lm(y_train ~ index + index2 + index3 + month + TEMP + DEWP + PRES + Iws + cbwdNE + cbwdNW + cbwdSE + cbwdcv, data=train)
summary(full_model)
```


Model diagnostics:

```{r}
library(lmtest)

par(mfrow=c(2,2))
plot(full_model)

bptest(full_model)
dwtest(full_model)
```

Ridge regression
```{r}
library(MASS)

stepwise_model <- stepAIC(full_model, direction="both", trace=FALSE)
summary(stepwise_model)
```


### interactions

```{r}
# interactions_model <- lm(y_train ~ index + index2 + index3 + month + TEMP + DEWP + PRES + cbwd + Iws + Is + Ir +
#                         TEMP:DEWP + TEMP:Iws + Iws:cbwd + TEMP:Ir + DEWP:Ir, data = train)
interactions_model <- lm(y_train ~ month + DEWP + Iws + TEMP - 1, data = train)
summary(interactions_model)
```


## Evaluate models

```{r}
rmse = function(actuals, predictions) {
  sqrt(mean((predictions - actuals)^2))
}
mape = function(actuals, predictions) {
  mean(abs((predictions - actuals) / actuals)) * 100
}
mae = function(actuals, predictions) {
  mean(abs(predictions - actuals))
}
```

```{r}
# Predictions
predictions_simple = predict(model, newdata=test)
predictions_full = predict(full_model, newdata=test)
predictions_stepwise = predict(stepwise_model, newdata=test)
predictions_interactions <- predict(interactions_model, newdata=test)


# Evaluate models
cat("Simple polynomial model Evaluation:\n")
cat("RMSE:", rmse(y_test, predictions_simple), "\n")
cat("MAPE:", mape(y_test, predictions_simple), "\n")
cat("MAE:", mae(y_test, predictions_simple), "\n\n")

cat("Full Model Evaluation:\n")
cat("RMSE:", rmse(y_test, predictions_full), "\n")
cat("MAPE:", mape(y_test, predictions_full), "\n")
cat("MAE:", mae(y_test, predictions_full), "\n\n")

cat("Stepwise Model Evaluation:\n")
cat("RMSE:", rmse(y_test, predictions_stepwise), "\n")
cat("MAPE:", mape(y_test, predictions_stepwise), "\n")
cat("MAE:", mae(y_test, predictions_stepwise), "\n\n")

cat("Interactions Model Evaluation:\n")
cat("RMSE:", rmse(y_test, predictions_interactions), "\n")
cat("MAPE:", mape(y_test, predictions_interactions), "\n")
cat("MAE:", mae(y_test, predictions_interactions), "\n\n")
```

### short term metrics

```{r}
y_test_subset = y_test[1:7]
predictions_simple_subset = predictions_simple[1:7]
predictions_full_subset = predictions_full[1:7]
predictions_stepwise_subset = predictions_stepwise[1:7]
predictions_interactions_subset = predictions_interactions[1:7]

cat("Simple polynomial model Evaluation for the first 7 data points:\n")
cat("RMSE:", rmse(y_test_subset, predictions_simple_subset), "\n")
cat("MAPE:", mape(y_test_subset, predictions_simple_subset), "\n")
cat("MAE:", mae(y_test_subset, predictions_simple_subset), "\n\n")

cat("Full Model Evaluation for the first 7 data points:\n")
cat("RMSE:", rmse(y_test_subset, predictions_full_subset), "\n")
cat("MAPE:", mape(y_test_subset, predictions_full_subset), "\n")
cat("MAE:", mae(y_test_subset, predictions_full_subset), "\n\n")

cat("Stepwise Model Evaluation for the first 7 data points:\n")
cat("RMSE:", rmse(y_test_subset, predictions_stepwise_subset), "\n")
cat("MAPE:", mape(y_test_subset, predictions_stepwise_subset), "\n")
cat("MAE:", mae(y_test_subset, predictions_stepwise_subset), "\n\n")

cat("Interactions Model Evaluation for the first 7 data points:\n")
cat("RMSE:", rmse(y_test_subset, predictions_interactions_subset), "\n")
cat("MAPE:", mape(y_test_subset, predictions_interactions_subset), "\n")
cat("MAE:", mae(y_test_subset, predictions_interactions_subset), "\n\n")
```


## Residuals plots

```{r}
residuals_simple <- y_test - predictions_simple
residuals_full <- y_test - predictions_full
residuals_stepwise <- y_test - predictions_stepwise
residuals_interactions <- y_test - predictions_interactions
residuals_ridge <- y_test - predictions_ridge

# par(mfrow=c(2,1)) 

plot(residuals_simple, main="Residuals for Baseline Model", ylab="Residuals", xlab="Time", pch=20)
abline(h=0, col="red")
plot(residuals_ridge, main="Residuals for Ridge Model", ylab="Residuals", xlab="Time", pch=20)
abline(h=0, col="red")
# plot(residuals_stepwise, main="Residuals for Stepwise Model", ylab="Residuals", xlab="Index", col='orange', pch=20)
# abline(h=0, col="red")
# plot(residuals_interactions, main="Residuals for Interactions Model", ylab="Residuals", xlab="Index", col='purple', pch=20)
# abline(h=0, col="red")
```

```{r}
library(glmnet)

x <- model.matrix(~ month + DEWP + Iws + TEMP - 1, data = train)
y <- y_train

ridge_model <- glmnet(x, y, alpha = 0)
predictions_ridge <- predict(ridge_model, s=0.1, newx = model.matrix(~ month + DEWP + Iws + TEMP - 1, data = test))
predictions_ridge_subset <- predictions_ridge[1:7]

cat("Ridge Regression Model Evaluation:\n")
cat("RMSE:", rmse(y_test, predictions_ridge), "\n")
cat("MAPE:", mape(y_test, predictions_ridge), "\n")
cat("MAE:", mae(y_test, predictions_ridge), "\n\n")

cat("Ridge Regression Model Evaluation for the first 7 data points:\n")
cat("RMSE:", rmse(y_test_subset, predictions_ridge_subset), "\n")
cat("MAPE:", mape(y_test_subset, predictions_ridge_subset), "\n")
cat("MAE:", mae(y_test_subset, predictions_ridge_subset), "\n\n")
```



